---
title: "Interactive Pharmacokinetic Modeling Exercise"
format: html
editor: visual
execute:
  shiny: true
  shinylive: true
---

## Introduction to Pharmacokinetic Modeling

Pharmacokinetic (PK) modeling is essential for understanding drug behavior in the body and optimizing dosing regimens. This interactive module allows you to explore how different patient characteristics and dosing parameters affect drug concentrations over time.

### Key Learning Objectives

- Understand the relationship between PK parameters and concentration-time profiles
- Explore the impact of different routes of administration
- Analyze steady-state achievement and accumulation
- Apply pharmacodynamic principles to optimize therapeutic outcomes
- Practice therapeutic drug monitoring concepts

### Before You Begin

This simulation uses a one-compartment pharmacokinetic model with first-order elimination. While simplified, this model captures the essential features of many drugs and provides excellent educational value for understanding fundamental PK/PD principles.

## Interactive Pharmacokinetic Modeling Module

```{r setup, include=FALSE}
library(shiny)
library(ggplot2)
library(dplyr)
library(plotly)
library(DT)
library(stringr)
```

::: {.panel-tabset}

### PK/PD Simulation

```{r}
ui <- fluidPage(
  titlePanel("Advanced Pharmacokinetic/Pharmacodynamic Modeling Module"),
  
  sidebarLayout(
    sidebarPanel(
      width = 4,
      
      # Patient Generation
      h4("Patient Characteristics"),
      actionButton("new_patient", "Generate Random Patient", class = "btn-primary"),
      br(), br(),
      
      # Dosing Parameters
      h4("Dosing Regimen"),
      selectInput("route", "Route of Administration:", 
                  choices = list("Oral" = "oral", "IV Bolus" = "iv_bolus", "IV Infusion" = "iv_infusion"),
                  selected = "oral"),
      numericInput("dose", "Dose (mg):", value = 500, min = 50, max = 2000, step = 50),
      conditionalPanel(
        condition = "input.route == 'iv_infusion'",
        numericInput("infusion_time", "Infusion Duration (hr):", value = 1, min = 0.25, max = 24, step = 0.25)
      ),
      numericInput("tau", "Dosing Interval (hr):", value = 12, min = 6, max = 48),
      numericInput("n_doses", "Number of Doses:", value = 10, min = 1, max = 30),
      
      hr(),
      
      # PK Parameters (Advanced)
      h4("PK Parameters (Advanced)"),
      checkboxInput("advanced_pk", "Enable Manual PK Parameter Control", FALSE),
      conditionalPanel(
        condition = "input.advanced_pk == true",
        numericInput("manual_CL", "Clearance (L/hr):", value = 6, min = 1, max = 20, step = 0.1),
        numericInput("manual_Vd", "Volume of Distribution (L):", value = 50, min = 10, max = 100, step = 1),
        numericInput("manual_ka", "Absorption Rate Constant (1/hr):", value = 1, min = 0.1, max = 5, step = 0.1),
        numericInput("manual_F", "Bioavailability (fraction):", value = 0.8, min = 0.1, max = 1, step = 0.05)
      ),
      
      hr(),
      
      # Patient Info Display
      h4("Current Patient:"),
      verbatimTextOutput("patient_info"),
      
      hr(),
      
      # PD Parameters
      h4("Pharmacodynamic Modeling"),
      checkboxInput("enable_pd", "Enable PD Simulation", FALSE),
      conditionalPanel(
        condition = "input.enable_pd == true",
        numericInput("ec50", "EC50 (mg/L):", value = 2, min = 0.1, max = 10, step = 0.1),
        numericInput("emax", "Emax (% effect):", value = 100, min = 10, max = 100, step = 5),
        numericInput("mic", "MIC (mg/L):", value = 1, min = 0.1, max = 8, step = 0.1)
      )
    ),
    
    mainPanel(
      width = 8,
      tabsetPanel(
        tabPanel("Concentration-Time Plot", 
                 plotlyOutput("pk_plot", height = "500px"),
                 br(),
                 h5("Plot Interpretation:"),
                 textOutput("plot_interpretation")
        ),
        tabPanel("PK/PD Analysis",
                 conditionalPanel(
                   condition = "input.enable_pd == true",
                   plotlyOutput("pd_plot", height = "400px"),
                   br(),
                   h5("PK/PD Metrics:"),
                   tableOutput("pkpd_metrics")
                 ),
                 conditionalPanel(
                   condition = "input.enable_pd == false",
                   div(style = "text-align: center; margin-top: 100px;",
                       h4("Enable PD Simulation to view pharmacodynamic analysis"),
                       p("Check the 'Enable PD Simulation' option in the sidebar to access PK/PD modeling features.")
                   )
                 )
        ),
        tabPanel("Simulation Data", 
                 h4("Complete Simulation Dataset"),
                 DT::dataTableOutput("sim_data_table")
        ),
        tabPanel("Parameter Explanations",
                 div(
                   h4("Pharmacokinetic Parameters"),
                   tags$ul(
                     tags$li(strong("Clearance (CL):"), " The volume of plasma cleared of drug per unit time (L/hr)"),
                     tags$li(strong("Volume of Distribution (Vd):"), " Apparent volume in which the drug is distributed (L)"),
                     tags$li(strong("Absorption Rate Constant (ka):"), " Rate at which drug is absorbed from dosing site (1/hr)"),
                     tags$li(strong("Bioavailability (F):"), " Fraction of administered dose that reaches systemic circulation"),
                     tags$li(strong("Half-life (t½):"), " Time required for drug concentration to decrease by 50%"),
                     tags$li(strong("Elimination Rate Constant (ke):"), " Rate at which drug is eliminated from the body (1/hr)")
                   ),
                   br(),
                   h4("Pharmacodynamic Parameters"),
                   tags$ul(
                     tags$li(strong("EC50:"), " Concentration producing 50% of maximum pharmacological effect"),
                     tags$li(strong("Emax:"), " Maximum achievable pharmacological effect"),
                     tags$li(strong("MIC:"), " Minimum Inhibitory Concentration - lowest concentration that inhibits bacterial growth")
                   ),
                   br(),
                   h4("Clinical Relevance"),
                   p("This simulation helps understand how different dosing regimens achieve therapeutic concentrations while minimizing toxicity. 
                     Key considerations include:"),
                   tags$ul(
                     tags$li("Steady-state achievement (typically after 5 half-lives)"),
                     tags$li("Peak and trough concentrations for therapeutic drug monitoring"),
                     tags$li("Time above MIC for antimicrobial efficacy"),
                     tags$li("AUC/MIC ratio for concentration-dependent killing")
                   )
                 )
        )
      )
    )
  )
)

server <- function(input, output, session) {
  # Generate a random patient with realistic PK parameters
  patient <- reactiveVal()
  
  observeEvent(input$new_patient, {
    # Generate correlated PK parameters based on patient characteristics
    weight <- round(runif(1, 60, 100), 1)
    age <- sample(18:80, 1)
    
    # Weight-based scaling for CL and Vd with some variability
    CL_base <- 0.08 * weight * runif(1, 0.8, 1.2)  # Typical: 0.08 L/hr/kg
    Vd_base <- 0.6 * weight * runif(1, 0.7, 1.3)   # Typical: 0.6 L/kg
    
    # Age effects on clearance (reduced in elderly)
    if (age > 65) {
      CL_base <- CL_base * runif(1, 0.7, 0.9)
    }
    
    patient(list(
      weight = weight,
      age = age,
      CL = round(CL_base, 2),
      Vd = round(Vd_base, 2),
      ka = round(runif(1, 0.5, 2.5), 2),
      F = round(runif(1, 0.6, 0.95), 2),
      ke = round(CL_base / Vd_base, 3),
      half_life = round(0.693 * Vd_base / CL_base, 1)
    ))
  }, ignoreInit = FALSE)

  # Get current PK parameters (either from patient or manual inputs)
  current_pk <- reactive({
    p <- patient()
    if (is.null(p)) return(NULL)
    
    if (input$advanced_pk) {
      list(
        CL = input$manual_CL,
        Vd = input$manual_Vd,
        ka = input$manual_ka,
        F = input$manual_F,
        ke = input$manual_CL / input$manual_Vd,
        half_life = 0.693 * input$manual_Vd / input$manual_CL,
        weight = p$weight,
        age = p$age
      )
    } else {
      p
    }
  })

  output$patient_info <- renderPrint({
    p <- current_pk()
    if (!is.null(p)) {
      cat(
        paste0(
          "Weight: ", p$weight, " kg\n",
          "Age: ", p$age, " years\n",
          "CL: ", p$CL, " L/hr\n",
          "Vd: ", p$Vd, " L\n",
          "ka: ", p$ka, " 1/hr\n",
          "F: ", p$F, "\n",
          "ke: ", round(p$ke, 3), " 1/hr\n",
          "Half-life: ", round(p$half_life, 1), " hr"
        )
      )
    }
  })

  # Enhanced simulation data with multiple routes
  sim_data <- reactive({
    req(current_pk())
    p <- current_pk()
    
    dose <- input$dose
    tau <- input$tau
    n_doses <- input$n_doses
    route <- input$route
    
    CL <- p$CL
    Vd <- p$Vd
    ka <- p$ka
    F <- p$F
    ke <- p$ke
    
    # Time grid: simulate up to last dose + 3 half-lives
    tmax <- tau * (n_doses-1) + 3 * p$half_life
    time <- seq(0, tmax, by = 0.1)
    conc <- rep(0, length(time))
    
    # Route-specific calculations
    if (route == "oral") {
      # Oral multiple dosing with first-order absorption
      for (i in 0:(n_doses-1)) {
        t_dose <- time - i*tau
        idx <- which(t_dose >= 0)
        conc[idx] <- conc[idx] + 
          (F*dose*ka)/(Vd*(ka-ke)) * (exp(-ke*t_dose[idx]) - exp(-ka*t_dose[idx]))
      }
    } else if (route == "iv_bolus") {
      # IV bolus multiple dosing
      for (i in 0:(n_doses-1)) {
        t_dose <- time - i*tau
        idx <- which(t_dose >= 0)
        conc[idx] <- conc[idx] + (dose/Vd) * exp(-ke*t_dose[idx])
      }
    } else if (route == "iv_infusion") {
      # IV infusion multiple dosing
      tinf <- input$infusion_time
      for (i in 0:(n_doses-1)) {
        t_dose <- time - i*tau
        idx_during <- which(t_dose >= 0 & t_dose <= tinf)
        idx_after <- which(t_dose > tinf)
        
        # During infusion
        if (length(idx_during) > 0) {
          conc[idx_during] <- conc[idx_during] + 
            (dose/(CL*tinf)) * (1 - exp(-ke*t_dose[idx_during]))
        }
        
        # After infusion
        if (length(idx_after) > 0) {
          conc[idx_after] <- conc[idx_after] + 
            (dose/(CL*tinf)) * (1 - exp(-ke*tinf)) * exp(-ke*(t_dose[idx_after]-tinf))
        }
      }
    }
    
    # Calculate PD effect if enabled
    if (input$enable_pd) {
      ec50 <- input$ec50
      emax <- input$emax
      effect <- emax * conc / (ec50 + conc)  # Emax model
    } else {
      effect <- rep(NA, length(conc))
    }
    
    tibble(
      time = time,
      concentration = pmax(conc, 0),  # Ensure non-negative
      effect = effect,
      above_mic = ifelse(input$enable_pd, conc > input$mic, NA)
    )
  })

  # Enhanced concentration-time plot
  output$pk_plot <- renderPlotly({
    sim <- sim_data()
    p <- plot_ly(sim, x = ~time, y = ~concentration, type = 'scatter', mode = 'lines',
                 line = list(color = 'blue', width = 2),
                 name = 'Concentration') %>%
      layout(
        title = paste("Pharmacokinetic Profile:", str_to_title(input$route), "Administration"),
        xaxis = list(title = "Time (hr)"),
        yaxis = list(title = "Concentration (mg/L)"),
        showlegend = TRUE
      )
    
    # Add MIC line if PD is enabled
    if (input$enable_pd) {
      p <- p %>% add_trace(y = rep(input$mic, nrow(sim)), type = 'scatter', mode = 'lines',
                          line = list(color = 'red', dash = 'dash', width = 2),
                          name = 'MIC')
    }
    
    p
  })

  # Pharmacodynamic plot
  output$pd_plot <- renderPlotly({
    if (!input$enable_pd) return(NULL)
    
    sim <- sim_data()
    
    plot_ly(sim, x = ~time, y = ~effect, type = 'scatter', mode = 'lines',
            line = list(color = 'green', width = 2),
            name = 'Pharmacological Effect') %>%
      layout(
        title = "Pharmacodynamic Effect Over Time",
        xaxis = list(title = "Time (hr)"),
        yaxis = list(title = "Effect (%)", range = c(0, 100)),
        showlegend = FALSE
      )
  })

  # PK/PD metrics table
  output$pkpd_metrics <- renderTable({
    if (!input$enable_pd) return(NULL)
    
    sim <- sim_data()
    p <- current_pk()
    
    # Calculate key metrics
    cmax <- max(sim$concentration, na.rm = TRUE)
    tmax <- sim$time[which.max(sim$concentration)]
    auc <- sum(diff(sim$time) * (sim$concentration[-1] + sim$concentration[-nrow(sim)])) / 2
    time_above_mic <- sum(sim$above_mic, na.rm = TRUE) * 0.1  # time step = 0.1 hr
    
    metrics <- data.frame(
      Metric = c("Cmax (mg/L)", "Tmax (hr)", "AUC (mg⋅hr/L)", 
                 "Time > MIC (hr)", "AUC/MIC ratio", "Half-life (hr)"),
      Value = c(
        round(cmax, 2),
        round(tmax, 1),
        round(auc, 1),
        round(time_above_mic, 1),
        round(auc / input$mic, 1),
        round(p$half_life, 1)
      ),
      stringsAsFactors = FALSE
    )
    
    metrics
  }, striped = TRUE, hover = TRUE)

  # Complete simulation data table
  output$sim_data_table <- DT::renderDataTable({
    sim <- sim_data()
    DT::datatable(
      sim %>% 
        mutate(
          time = round(time, 1),
          concentration = round(concentration, 3),
          effect = round(effect, 1)
        ) %>%
        filter(row_number() %% 10 == 1),  # Show every 10th row for performance
      options = list(pageLength = 15, scrollX = TRUE)
    )
  })

  # Plot interpretation
  output$plot_interpretation <- renderText({
    p <- current_pk()
    if (is.null(p)) return("")
    
    route_text <- switch(input$route,
      "oral" = "oral absorption phase followed by elimination",
      "iv_bolus" = "immediate peak followed by first-order elimination", 
      "iv_infusion" = "gradual increase during infusion, then elimination"
    )
    
    paste("This plot shows the", route_text, "pattern. The half-life is", 
          round(p$half_life, 1), "hours, and steady-state is typically achieved after",
          round(5 * p$half_life, 1), "hours of repeated dosing.")
  })
}

shinyApp(ui, server)
```
:::

## Guided Learning Exercises

Use the interactive module above to complete these exercises:

### Exercise 1: Route of Administration Comparison
1. Generate a random patient and note their PK parameters
2. Set up a dosing regimen: 500 mg every 12 hours for 5 doses
3. Compare the concentration-time profiles for:
   - Oral administration
   - IV bolus
   - IV infusion (1-hour duration)
4. **Questions:**
   - Which route achieves the highest peak concentration?
   - Which route shows the most gradual concentration changes?
   - How does bioavailability affect the oral vs IV comparison?

### Exercise 2: Steady-State Analysis
1. Generate a new patient
2. Use oral dosing: 250 mg every 8 hours for 15 doses
3. Observe the concentration-time profile
4. **Questions:**
   - At approximately what time is steady-state achieved?
   - How does this relate to the patient's half-life?
   - What happens to peak and trough concentrations over time?

### Exercise 3: Pharmacodynamic Modeling
1. Enable PD simulation with default parameters (EC50 = 2 mg/L, MIC = 1 mg/L)
2. Try different dosing regimens to optimize:
   - Time above MIC
   - AUC/MIC ratio
3. **Questions:**
   - Which metric is more important for antimicrobial efficacy?
   - How do you balance efficacy with potential toxicity?

### Exercise 4: Patient Variability
1. Generate multiple patients and compare their PK parameters
2. Use the same dosing regimen for different patients
3. Enable advanced PK controls and manually adjust parameters
4. **Questions:**
   - How does clearance variability affect dosing requirements?
   - Why might elderly patients require dose adjustments?
   - What role does body weight play in PK parameter scaling?

### Key Takeaways

- **Steady-state** is achieved after approximately 5 half-lives
- **Route of administration** significantly impacts concentration-time profiles
- **Patient characteristics** (age, weight) influence PK parameters
- **PK/PD modeling** helps optimize dosing for both efficacy and safety
- **Therapeutic drug monitoring** uses peak and trough measurements to guide dosing





::::: {.d-flex .justify-content-between .border-top .pt-4 .mt-4}
<div>

[← Session 1](index.qmd){.btn .btn-outline-primary}

</div>

<div>

[Session 2 →](../session_02/index.qmd){.btn .btn-outline-primary}

</div>
:::::
